---
title: "Species distribution modelling with Poisson point processes"
subtitle: "The Quantitative & Applied Ecology Group (QAECO), The University of Melbourne"
author: "Ian Flint and Roozbeh Valavi"
date: "`r Sys.Date()`"
output: 
  pdf_document
vignette: >
  %\VignetteIndexEntry{Species distribution modelling with Poisson point processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, tidy.opts=list(width.cutoff=80), fig.align="center", message = FALSE, warning = FALSE, fig.height=5, fig.width=5)

set.seed(42)
```

## Species distribution modelling

Species distribution modelling is...


The example data used in this tutorial comes form Fithian and others (2015).

## Envrionmental data

The environmental data are a series of climatic and topographic variables. We already transferred the coordinate reference system (CRS) of all the input data from geographic CRS to a metric system (UTM zone 56, south). So, the unit of the maps is metre.


```{r}
# library(maptools)
library(raster)

grid_dir <- "data/grids_utm"
vars <- list.files(grid_dir, pattern = ".tif$", full.names = TRUE)

# read the raster layers as a raster stack
nsw_stack <- stack(vars)

# you can plot them by:
# plot(nsw_stack)

```

## Species data

The species presence-only data consists of 36 eucalypt species in south-eastern Australia. We use an example species to fit `MaxNet` and `PPM` models. As documented in Fithian et al. (2015), the species data has a bias towards cities and towns, especially toward Sydney, the biggest city in the study area. The rest of the species is used to generate Target Group Background (TGB) samples to be used as one way to account for the sampling bias in the presence-only species data in MaxNet.

```{r}
# reading species occurrence data
occ <- read.csv("data/species/eucablax.csv")
str(occ)

# plot the sapecies on the raster map
plot(nsw_stack[[1]])
points(occ$X, occ$Y)
```


## Modelling with Maxent

Maxent is a very popular tool for modelling species occurrence data. Maxent was first developed as a stand-alone Java programme and later a wrapper function was implemented in the `dismo` package to fit SDMs with Maxent in R programming.
Here we use `maxnet` package, the open-source version of `Maxent` in R, newly developed by Phillips *et al.,* (2017) to model species distributions from occurrences and environmental variables, using `glmnet` for model fitting.


```{r}
# load the libraries
library(maxnet)
library(dismo) # for generating random points

```

For model fitting with maxnet, we need a vector of presence and background data (with 0 indicating backgrounds and 1 indicating presences) and a `data.frame` of the environmental covariates.
    
First, we need to get the background samples. The background samples are typically selected randomly from the landscape. The number of samples should high enough to represent the environmental variability in the landscape of interest. Based on the convention, 10,000 random background is commonly used for model fitting with background, However, evidence were provided in the literature related to spatial point processes that this number is not enough. Here we have `r length(which(!is.na(values(nsw_stack[[1]]))))` number of non-NA pixels in our landscape, so clearly the 10,000 is not enough. Although, for the purpose of this tutorial, we limit the number of background here to speeds up the computation.

After generating the random bacground samples, we need to extract the values of environmental variables. This will be a `data.frame` that later will be used in model fitting.


```{r}
# generate some random samples
rnd <- randomPoints(nsw_stack, n = 25000)

# extracting the values of rasters for the background points
background <- raster::extract(nsw_stack, rnd)
head(background)

# check for any NA in the extracted values
anyNA(background)

# remove the NA values
background <- na.omit(background)

```

The values of the environmental variable for the species occurrences also need to be extracted. Then the covariates of the presence and background should be combined in a single `data.frame`.

```{r}
# subset the covariate in species data by raster
occ_env <- raster::extract(nsw_stack, occ[,c("X", "Y")])
# now add them together
covariates <- as.data.frame(rbind(occ_env, background))
head(covariates)
```

Here, we need to make a voctor of 1s and 0s. This should be with the same order as the covariate data is provided.

```{r fig.height=6.5, fig.width=7}
# make the presence and background points
# presnece (1s) and background (0s)
# the order should be the same as the order in the covariates
presences <- c(rep(1, nrow(occ)), 
               rep(0, nrow(background)))

tmp <- Sys.time()
mxnet <- maxnet(p = presences,
                data = covariates,
                regmult = 1, # regularisation multiplier
                maxnet.formula(presences, covariates, classes = "lqph"))
Sys.time() - tmp

# plot the fitted function
# plot(mxnet, type = "cloglog")

```

### Pridicting on rasters

To predict this model on rasters, you need a `RasterStack` with all the covariates used in model fitting. The names of the covariates in the `RasterStack` must be exactly the same. This is done through `raster` package (already loaded with `dismo` pcakage).

There are different options for scaling the raw output of `maxnet` including the **link**, **exponential**, **logistic** and recently introduced **cloglog**. For more information about these option please read Phillips et al. (2017). To show a progress bar use `progress = "text"`.

```{r}
tmp <- Sys.time()
max_pred <- raster::predict(object = nsw_stack, model = mxnet, type = "cloglog")
Sys.time() - tmp

# plot the prediction
plot(max_pred)

```

### Bias Correction

Creating a map of distance to the towns to correct for biases in the data.

```{r}
library(sf)

# the border shapefile
border <- st_read("data/border/ibraone.shp", quiet = TRUE)
# reading the town point data
towns <- st_read("data/towns/towns.shp")

plot(nsw_stack[[1]], main = "Towns in the region")
plot(st_geometry(towns), add = TRUE)

```

```{r}
distmap <- distanceFromPoints(nsw_stack[[1]], towns) # this may take some time
distmap <- mask(distmap, mask = border) # crop base on the region

plot(distmap)
points(occ$X, occ$Y)
```



## Modelling with point processes

The `spatstat` package is the most widely used to work with point processes.
Covariates are usually specified in their image objects `spatstat::im`.
Internally, this is represented as a large pixel matrix, so conversion from rasters and other image objects is usually straightforward.

```{r}
library(spatstat)

covariates <- lapply(as.list(nsw_stack), function(element) maptools::as.im.RasterLayer(element))
names(covariates) <- names(nsw_stack)
```

Spatstat also needs to be told what the observation region is.
The required object type is `spatstat::owin`.
Common ways to construct an `owin` is to either take a fixed rectangle, i.e. `window <- owin(c(0, 100), c(0, 100))`, or to use an existing covariate or raster to construct the window.
The latter technique is what we will use here.

Although it would be possible to do `window <- spatstat::as.owin(covariates[[1]])`, it will be easier to work on a window with a lower resolution, as shown next.

```{r}
window <- spatstat::as.owin(as.mask(covariates[[1]], eps = 100))
```

Locations of individuals are represented via a point pattern object `spatstat::ppp`, and consist in coordinates along with a window in which the species has been observed.

```{r}
configuration <- spatstat::ppp(x = occ$X, y = occ$Y, window = window)
```

Point patterns can easily be plotted.

```{r}
plot(configuration)
```

It is usually a good idea to start by a ``static'' analysis of the point pattern, without yet involving covariates.

```{r}
summary(configuration)
plot(spatstat::density.ppp(configuration))
plot(spatstat::Kest(configuration))
# The line below takes 3 min to execute and is not crucial to the analysis.
# plot(spatstat::envelope(configuration,Kest))
```

Doing inference on the point pattern is just as easy as setting up a `glm` regression.
Start by writing the formula, essentially `formula <- "configuration ~ covariates`

```{r}
formula <- paste0("configuration ~ ", paste0(names(covariates), collapse = " + "))
print(formula)
```

The fitting function (analogue of `glm`) is `spatstat::ppm` and is used as follows.

```{r}
fit <- spatstat::ppm(as.formula(formula), covariates = covariates)
```

The fitted regression is manipulated in the same way as a `glm` fit is, so for example you can have a look at the summary

```{r}
summary(fit)
```

or do an ANOVA.

```{r}
formula_without_bc04 <- paste0("configuration ~ ", paste0(names(covariates)[-2], collapse = " + "))
fit_without_bc04 <- spatstat::ppm(as.formula(formula_without_bc04), covariates = covariates)
anova(fit, fit_without_bc04)
```

To look at the predicted intensity, you use the `spatstat::predict.ppm` function.

```{r}
pred <- spatstat::predict.ppm(fit, covariates = covariates)
plot(pred)
```

Spatstat can handle many different types of correlation structures between individuals of the species.
You would usually supply an `interaction` parameter to `spatstat::ppm`.
However, initial analysis suggested attraction between the individuals, in which case a doubly-stochastic (Cox) point process is more appropriate.
Fitting such point processes uses another function, as shown below.

```{r}
fit_cox <- spatstat::kppm(as.formula(formula), covariates = covariates, clusters = "LGCP")
summary(fit_cox)
```

A nice way to appreciate the difference in the underlying model is to draw from the fitted distribution.
This can easily be done for the fitted Poisson point process.

```{r}
draw_ppp <- spatstat::simulate.ppm(fit)
plot(draw_ppp)
```

Drawing from a Cox point process requires you to use another library, but it essentially works in the same way.

```{r}
library(RandomFields)
library(RandomFieldsUtils)

draw_cox <- spatstat::simulate.kppm(fit_cox)
plot(draw_cox)
```

Making goodness-of-fit tests is straightforward, we refer in particular to the functions `spatstat::quadrat.test`, `spatstat::cdf.test`, `spatstat::dclf.test` and `spatstat::mad.test`.
A lot of these functions rely on multiple simulations of the point process, which is going to be exeedingly slow for the Cox process.
Instead, we show what a goodness-of-fit test looks like with a simple fit with a Poisson point process.

```{r}
dclf.test(fit)
```

## Comparing the results


