---
title: "Tutorial 1: simplest point process model"
author: "Ian Flint and Roozbeh Valavi"
date: "`r Sys.Date()`"
output: 
  pdf_document
vignette: >
  %\VignetteIndexEntry{Species distribution modelling with Poisson point processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
always_allow_html: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, tidy.opts = list(width.cutoff = 80), fig.align="center", message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5)

set.seed(1)
```

## Setup and R libraries

```{r}
## working with spatial data
library(maptools) # convert raster to spatstat format
library(terra) # working with raster data

## modelling 
library(spatstat) # fitting and exploring ppm models
```


```{r}
tmap::tmap_mode("view") # setting for tmap package

```


## Species data

Reading species data....

```{r}
occurrences <- read.csv("data/species/occurrences.csv")
head(occurrences)

```


## Environmental data

```{r}
files <- list.files("data/rasters/", pattern = ".tif$", full.names = TRUE)
# get the 19 bio-climatic variables
rasters <- terra::rast(files)

# plot a few of them
plot(rasters[[1]])
points(occurrences)

```


## Converting everything to `spatstat` objects

Covariates are usually specified in their image objects `spatstat::im`.
Internally, this is represented as a large pixel matrix, so conversion from rasters and other image objects is usually straightforward.
In order to convert to the proper format, the `maptools` helper function is useful.

```{r}
covariates <- lapply(rasters, as.im)
names(covariates) <- names(rasters)

plot(covariates[[1]], main = names(covariates)[1])

```


Next, when working with `spatstat` you need an observation window (region).
It is a region in which the points are assumed to be drawn from. 
The probability of finding locations outside of this region is assumed to be zero.
Common ways to construct an `spatstat::owin` is to either take a fixed rectangle, i.e. `window <- owin(c(0, 100), c(0, 100))`, or to use an existing covariate or raster to construct the window.

```{r}
window <- spatstat.geom::as.owin(covariates[[1]])
plot(window)

```


Occurrences (or presence points) are `spatstat::ppp` objects, and basically you need a vector of x-values, a vector of y-values, and the observation window.

```{r}
configuration <- spatstat.geom::ppp(x = occurrences$X, 
                                    y = occurrences$Y,
                                    window = window)
plot(configuration)

```

## Modelling with point processes

It is usually a good idea to start by a ``static'' analysis of the point pattern, without yet involving covariates.

```{r}
summary(configuration)
plot(spatstat.core::density.ppp(configuration))
points(configuration, pch = 16, cex = 0.2)

```


```{r}
plot(spatstat.core::Kest(configuration))

```

Doing inference on the point pattern is just as easy as setting up a `glm` regression.
Start by writing the formula, essentially `formula <- "configuration ~ covariates`

```{r}
formula <- paste0("configuration ~ ", paste0(names(covariates), collapse = " + "))
print(formula)
```

The fitting function (analogue of `glm`) is `spatstat::ppm` and is used as follows.

```{r}
fit <- spatstat::ppm(as.formula(formula), covariates = covariates)
```

The fitted regression is manipulated in the same way as a `glm` fit is, so for example you can have a look at the summary

```{r}
summary(fit)
```

or do an ANOVA.

```{r}
formula_without_tmin <- paste0("configuration ~ ", paste0(names(covariates)[names(covariates) != "tmin"], collapse = " + "))
fit_without_tmin <- spatstat::ppm(as.formula(formula_without_tmin), covariates = covariates)
anova(fit, fit_without_tmin)
```

To look at the predicted intensity, you use the `spatstat::predict.ppm` function.

```{r}
pred <- spatstat::predict.ppm(fit, covariates = covariates, dimyx = c(1024, 1024))
plot(log(pred))
```

Spatstat can handle many different types of correlation structures between individuals of the species.
You would usually supply an `interaction` parameter to `spatstat::ppm`.
However, initial analysis suggested attraction between the individuals, in which case a doubly-stochastic (Cox) point process is more appropriate.
Fitting such point processes uses another function, as shown below.

<!-- ```{r} -->
  <!-- fit_cox <- spatstat::kppm(as.formula(formula), covariates = covariates, clusters = "LGCP") -->
  <!-- summary(fit_cox) -->
  <!-- ``` -->
  
  <!-- A nice way to appreciate the difference in the underlying model is to draw from the fitted distribution. -->
  <!-- This can easily be done for the fitted Poisson point process. -->
  
  <!-- ```{r} -->
  <!-- draw_ppp <- spatstat::simulate.ppm(fit) -->
  <!-- plot(draw_ppp) -->
  <!-- ``` -->
  
  <!-- Drawing from a Cox point process requires you to use another library, but it essentially works in the same way. -->
  
  <!-- ```{r} -->
  <!-- library(RandomFields) -->
  <!-- library(RandomFieldsUtils) -->
  
  <!-- draw_cox <- spatstat::simulate.kppm(fit_cox) -->
  <!-- plot(draw_cox) -->
  <!-- ``` -->
  
  <!-- Making goodness-of-fit tests is straightforward, we refer in particular to the functions `spatstat::quadrat.test`, `spatstat::cdf.test`, `spatstat::dclf.test` and `spatstat::mad.test`. -->
  <!-- A lot of these functions rely on multiple simulations of the point process, which is going to be exeedingly slow for the Cox process. -->
  <!-- Instead, we show what a goodness-of-fit test looks like with a simple fit with a Poisson point process. -->
  
  <!-- ```{r} -->
  <!-- dclf.test(fit) -->
  <!-- ``` -->
  
  <!-- ## Comparing the results -->
  
  
  