---
title: "Spatial point process modelling"
author: "Ian Flint and Roozbeh Valavi"
date: "`r Sys.Date()`"
output: 
  pdf_document
vignette: >
  %\VignetteIndexEntry{Species distribution modelling with Poisson point processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
always_allow_html: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center", 
                      message = FALSE, warning = FALSE, 
                      fig.height = 5, fig.width = 5)

set.seed(1)
```

In this document, we show how to fit the simplest point process model namely, Poisson process model.


# Simplest point process model

## Setup and R libraries

```{r}
library(spatstat) # fitting and exploring ppm models
library(terra) # working with raster data

```


```{r}
source("R/helper_functions.R")

```


## Species data

Reading species data....

Note that the spatial coordinate system is a metric CRS...

```{r}
occurrences <- read.csv("data/species/occurrences.csv")
head(occurrences)

```


## Environmental data

We read raster covariate here....

We need to centre and scale covariate to ... 

```{r}
# get the 19 bio-climatic variables
rasters <- terra::scale(terra::rast(list.files("data/rasters/", pattern = ".tif$", full.names = TRUE)),
                        center = TRUE, scale = TRUE)

# plot a few of them
plot(rasters[[1]])
points(occurrences)

```


## Converting everything to `spatstat` objects

Covariates are usually specified in their image objects `spatstat::im`.
Internally, this is represented as a large pixel matrix, so conversion from rasters and other image objects is usually straightforward.
In order to convert to the proper format, the `maptools` helper function is useful.


```{r}
covariates <- lapply(rasters, spatstat.geom::as.im)
names(covariates) <- names(rasters)

plot(covariates[[1]], main = names(covariates)[1])

```


Next, when working with `spatstat` you need an observation window (region).
It is a region in which the points are assumed to be drawn from. 
The probability of finding locations outside of this region is assumed to be zero.
Common ways to construct an `spatstat::owin` is to either take a fixed rectangle, i.e. `window <- owin(c(0, 100), c(0, 100))`, or to use an existing covariate or raster to construct the window.

```{r}
window <- spatstat.geom::as.owin(covariates[[1]])
plot(window)

```


Occurrences (or presence points) are `spatstat::ppp` objects, and basically you need a vector of x-values, a vector of y-values, and the observation window.

```{r}
configuration <- spatstat.geom::ppp(x = occurrences$X, 
                                    y = occurrences$Y,
                                    window = window)
plot(configuration)

```


## Model fitting

Doing inference on the point pattern is just as easy as setting up a `glm` regression.
Start by writing the formula, essentially `formula <- "configuration ~ covariates`.
In the formula, we use `polynom` to allow the user to easily specify whether to use higher order polynomials in the model.

```{r}
formula <- configuration ~ 1 + bio_4 + bio_5 + bio_12 + bio_15

```


The fitting function (analogue of `glm`) is `ppm` and is used as follows.

```{r}
fit <- spatstat.core::ppm(formula, covariates = covariates)

```


## Checking model

The fitted regression is manipulated in the same way as a `glm` fit is, so for example you can have a look at the summary

```{r}
summary(fit)

```

To look at the predicted intensity, you use the `spatstat.core::predict.ppm` function.

## Prediction maps

```{r}
pred <- spatstat.core::predict.ppm(fit, covariates = covariates, dimyx = c(1024, 1024))

```

```{r}
plot(log(rast(pred)))
points(configuration, pch = 16, cex = 0.2)
```


# Point process advanced topics

Requirements from the previous sections to run this section:
- `configuration` object form section x.1
- `fit` object from sections x.3

In this section we show some more ...

## Post prediction checks

Here, we use `K` function to assess the clustering on the residual of the model. There are other functions such as `L` function and `g` function.

<!-- 
```{r}
k_func <- spatstat.core::Kinhom(X = configuration,
                                lambda = pred,
                                correction = "translate", 
                                normpower = 2)

plot(k_func)

```
 -->


```{r}
k_func <- spatstat.core::envelope(Y = fit, 
                                  fun = spatstat.core::Kinhom, # Use the inhomogeneous K function
                                  correction = "translate", 
                                  normpower = 2) # Recommended option

plot(k_func)

```

There are clear and unambiguous signs of clustering at all distances, suggesting modelling with Cox processes.

## Cox process

Spatstat can handle many different types of correlation structures between individuals of the species.
You would usually supply an `interaction` parameter to `spatstat.core::ppm`.
However, initial analysis suggested attraction between the individuals, in which case a doubly-stochastic (Cox) point process is more appropriate.
Fitting such point processes uses another function, as shown below.

```{r}
fit_cox <- kppm(formula, covariates = covariates, clusters = "LGCP")
summary(fit_cox)

```

Simulating a Cox point process essentially works as in the Poisson setting.

```{r}
draw_cox <- simulate.kppm(fit_cox)
plot(draw_cox)

```

Making goodness-of-fit tests is straightforward, we refer in particular to the functions `spatstat::quadrat.test`, `spatstat::cdf.test`, `spatstat::dclf.test` and `spatstat::mad.test`.
A lot of these functions rely on multiple simulations of the point process.

```{r}
dclf.test(fit, nsim = 20)
dclf.test(fit_cox, nsim = 20)

k_func_cox <- spatstat.core::envelope(Y = fit_cox, 
                                      fun = spatstat.core::Kinhom, # Use the inhomogeneous K function
                                      correction = "translate", 
                                      nsim = 20,
                                      normpower = 2) # Recommended option

plot(k_func_cox)

```
  
  
## Other diagnostics

```{r}
quadrat.test(fit)
cdf.test(fit, covariate = "x")

ss <- simulate(fit, nsim = 5)
plot(ss)

plot(log(raster::raster(density(ss, sigma = 90000)[[1]])))
```

How to validate models? We can try AUC

```{r}
auc(configuration, pred)
```
  
  
## Pre-modelling diagnostics

It is usually a good idea to start by a `static` analysis of the point pattern, without yet involving covariates.

```{r}
summary(configuration)

point_density <- spatstat.core::density.ppp(configuration, sigma = 1e5)
plot(log(point_density + 1e-11)) # a very small arbitrary value to avoid log of zero
points(configuration, pch = 16, cex = 0.2)

```


```{r}
###*********** You can ignore this Ian **********######
library(sf)

# make a grid with the size ncol = 20
grids <- sf::st_make_grid(rasters, n = 20, square = FALSE, what = "polygon") %>% 
  sf::st_sf() %>% 
  dplyr::mutate(id = 1:nrow(.))

rpoints <- terra::spatSample(rasters[[1]],
                  size = 100000,
                  method= "regular",
                  na.rm = TRUE,
                  as.points = TRUE) %>%
  sf::st_as_sf()
grids <- grids[rpoints,]

# count the numbers in the grid
occ_sf <- sf::st_as_sf(occurrences, coords = 1:2, crs = 7845)
grids$n_occ = lengths(st_intersects(grids, occ_sf))

ggplot() +
  geom_sf(data = grids, aes(fill = log(n_occ)), alpha = 0.7) +
  scale_fill_distiller(palette = 1, direction = 1, na.value = "white")

```


## Model selection with point process models


or do an ANOVA.




```{r}
# some few updated models with one less covariates
fit1 <- spatstat.core::ppm(configuration ~ 1 + bio_4 + bio_5 + bio_12, covariates = covariates)
fit2 <- spatstat.core::ppm(configuration ~ 1 + bio_4 + bio_5 + bio_15, covariates = covariates)
fit3 <- spatstat.core::ppm(configuration ~ 1 + bio_4 + bio_12 + bio_15, covariates = covariates)
fit4 <- spatstat.core::ppm(configuration ~ 1 + bio_5 + bio_12 + bio_15, covariates = covariates)
```

The ANOVA with highest reduction shows the variable with the highest impact on the model.

```{r}
anova(fit, fit1)
anova(fit, fit2)
anova(fit, fit3)
anova(fit, fit4)

```



## Fit Poisson process in Stan (Bayesian framework)

[](https://towardsdatascience.com/understanding-point-process-model-with-r-983553ca2a86)

```{r}
ppm_stan <- '
data{
  int<lower = 1> n;
  vector[n] x;
  int<lower = 0> y[n];
}
parameters{
  real beta0;
  real beta1;
}
transformed parameters{
}
model{
  //priors
  target += normal_lpdf(beta0 | 0,5);
  target += normal_lpdf(beta1 | 0,10);// likelihood
  target += poisson_log_lpmf(y | beta0 + beta1 * x);
}
generated quantities{
  vector[n] lambda_rep;
  lambda_rep = exp(beta0 + beta1 * x);
}'

```






# Point process model evaluation

Requirements from the previous sections to run this section:
- `fit` object form section x.1
- `covars` object from sections x.3






