---
title: 'Tutorial 0: data preparation for point process modelling'
author: "Ian Flint and Roozbeh Valavi"
date: "`r Sys.Date()`"
output: 
  pdf_document
vignette: >
  %\VignetteIndexEntry{Species distribution modelling with Poisson point processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
always_allow_html: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, tidy.opts = list(width.cutoff = 80), fig.align="center", message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5)

set.seed(1)
```



## Setup and R libraries

```{r}
## working with spatial data
library(sf) # working with spatial vector data
library(terra) # working with raster data

## data cleaning and plotting
library(geodata) # downloading raster data for modelling
# remotes::install_github("ropensci/rnaturalearthhires") # you also need this
library(rnaturalearth)
library(spocc) # downloading example species data
library(tmap) # plotting spatial data
library(tidyverse) # data cleaning and wrangling

```


```{r}
tmap::tmap_mode("view") # setting for tmap package
# a folder to save raster temp data
download_path <- "outputs/"
# reading helper functions
source("R/helper_functions.R")

```




## Downloading species data

We begin by loading the data containing recorded occurrences of koalas.

```{r}
occ <- spocc::occ(query = "Phascolarctos cinereus", 
                  from = "gbif",
                  date = c("2010-01-01", "2020-12-31"),
                  gbifopts = list(country = "AU"),
                  has_coords = TRUE, 
                  limit = 5e3)

occ_df <- spocc::occ2df(occ)
head(occ_df)

```




The locations are shown below.

```{r}
occ_sf <- sf::st_as_sf(occ_df, coords = c("longitude", "latitude"), crs = 4326)

tmap::tm_basemap() +
  tmap::tm_shape(occ_sf) + 
  tmap::tm_dots()

```

Citizen science data often have duplicated, unexpected locations (e.g. location of an animal in a zoo far from it actual habitat), points of terrestrial animal and plants locating in ocean, mistake coordinates etc. We explore these in one of the following sections.


## Environmental covariates

```{r}
australia <- rnaturalearth::ne_countries(country = "Australia",
                                         scale = 50,
                                         returnclass = "sf") %>% 
  sf::st_crop(c(xmin = 110, xmax = 155, ymin = -45, ymax = -5))


koala_states <- c("South Australia",
                  "Queensland", 
                  "New South Wales", 
                  "Victoria", 
                  "Australian Capital Territory")
eastern_states <- rnaturalearth::ne_states(country = "Australia",
                                           returnclass = "sf") %>% 
  dplyr::filter(name %in% koala_states) %>% 
  sf::st_union() %>% 
  terra::vect()


# get the 19 bio-climatic variables
raster_covars <- geodata::worldclim_global(var = "bio",
                                           res = 10,
                                           path = download_path) %>% 
  terra::crop(eastern_states) %>% 
  terra::mask(eastern_states)


# cleaning the names of covariates (optional)
names(raster_covars) <- substr(names(raster_covars), 11, 100)

# plot a few of them
plot(raster_covars[[1:6]])

```

## Removing correlated covariates



We check collinearity among a few pre-selected covariate as these are know to be relevant to koala distribution

```{r}
covar_list <- c("bio_4", "bio_7", "bio_5", "bio_12", "bio_15")
vif <- usdm::vifstep(raster::stack(raster_covars[[covar_list]]))
vif

```


```{r}
# select the list of uncorrelated covariates
uncorr_list <- vif@results$Variables
selected_covars <- raster_covars[[as.character(uncorr_list)]]

```



## Let's clean the coordinates of occurrence


Cleaning the data to remove unlikely points... in this case there are some points in .... that we remove by filtering the longitude. You might need to check your desired species data and remove the unlikely records.

```{r}
occ_cleaned <- dplyr::filter(occ_df, longitude > 135.3516) %>% 
  dplyr::select(longitude, latitude)

head(occ_cleaned)
```


Make use that all the layers have the same NA data. By taking a `min` or `max` function a mask layer is created as an one raster layer with NA will result in NA in the mask layer.


```{r}
# function is from the helper_functions.R
occ_cleaned <- clean_occ(points = occ_cleaned, 
                         r = selected_covars[[1]],
                         crs = 4326,  # the default CRS of GBIF records;
                         remove_outside = TRUE, 
                         remove_dup_dist = 0)

```




## Data projection

For PPM it is better to have the data in a metric reference systems...

Some ramble on CRS by Roozbeh...

We need to change both raster and point data into one projections.

```{r}
# GDA 2020 Lambert CRS
output_crs <- 7845

```


```{r}
rasters <- terra::project(selected_covars, paste0("epsg:", output_crs))
plot(rasters[[1]], main = names(rasters)[1])

```


```{r}
occ_transformed <- crs_transform(occ_cleaned, 
                                 in_crs = 4326, 
                                 out_crs = output_crs)
head(occ_transformed)

```


## Writing data to disk

```{r}
write.csv(occ_transformed, "data/species/occurrences.csv", row.names = FALSE)

```


```{r}
terra::writeRaster(
  x = rasters,
  filename = paste0("data/rasters/", names(rasters), ".tif"),
  overwrite = TRUE
)

```








